Notes and code examples to accompany "If the Null Fits, You Must Omit

### 2025-11-24 GroupPolarizationStatmod/paper/method.qmd
finished expectation model subsection

---
editor: 
  markdown: 
    wrap: 100
---

@@

A group polarization replication is proven undecidable, and therefore not truly a replication, if we
can find a model of *mere agreement* that appears to be polarized under that replication's
experimental design. If we cannot find such a model, we cannot determine a replication to be
undecidable. Mere agreement in this model is a dynamic where agreement increases within the group,
but extremism does not---in other words, A *null-polarization distributions* is a set of two
Gaussian distributions: one representing the pre-discussion opinions of the group, and the other
representing the post-discussion opinions. Our *null-polarization distributions* represent the
process of the group finding *mere agreement*, which is when the average group opinion is constant
over time, but opinion variance decreases to represent agreement (or, equivalently,
*consensus* [@DeGroot1974]).

@@

## Simulation of Mere Consensus and Spurious Polarization

Our analysis models \emph{mere consensus} to examine whether published detections of group polarization may plausibly be \emph{spurious}, caused instead by artificats introduced by opinion measurement design.

Latent opinions are modeled as normally-distributed random variables. This is a simple, sufficiently
general way to represent group-level changes in opinions as encoded in the general hypothesis of
group polarization, that likeminded groups tend to become more extreme on average as they discuss.
For time, $t$, then, we represent each participant $i$'s opinion as 
$$
\omega_{i,t} \sim \mathcal{N}(\mu_t, \sigma_t).
$$

Group-level opinion change, then, can be represented as a change in opinion mean, opinion variance,
or both. Group polarization occurs when a group's average opinion increases over time. When
likeminded people interact, they tend to come to agree more over time---they *find consensus*. By
definition this means opinion variance decreases, whether the average changes or not.

We can define a null-polarization model, then, as one where the group's average opinion stays
constant, but the variance decreases: $\sigmapost
\leq \sigmapre$. Since the mean opinion is constant we just call it $\mu = \mupre = \mupost$. For
conciseness, we define a null-polarization model as a 3-tuple, $\nu = (\mu, \sigmapre, \sigmapost)$.

@@

### Simulated group polarization measurements

We simulate the collection of an *ordinal distribution* of *ordinal observations*, as was performed
in a prototypical group polarization experiment by integrating the latent opinion probability
density function, $f(\omega)$, over each bin's corresponding range of opinion values. For example,
consider an ordinal scale that runs from -3 (strongly disagree) to +3 (strongly agree) (as used by
@Moscovici1969, for example). the frequency with which the "2" bin is chosen is just the total area
underneath the probability density curve between 1.5 and 2.5.

Formally, the probability density function is $$
f(\omega) = 
    \frac{1}{\sigma \sqrt{2 \pi}} e^{-\frac{(\omega - \mu)^2}{2 \sigma^2}}.
$$ To set up our measurement simulation, we will represent oridnal bin values as $b_i$ with
$i=1,\ldots,B$---i.e., there are $B$ bins. To create $B$ bins there must be $B+1$ *thresholds* that
define what range of latent opinions correspond to which ordinal bin value.

@@

We model the measurement of ordinal opinions by calculating the probability density function over
bins, written $$
f[b_i] = \int_{\theta_{i - 1}}^{\theta_i} f(\omega) d\omega,
$$ switching to square brackets for the argument to $f$ now to emphasize that bins are discrete.
With this ordinal measurement model we can caluclate the mean opinion of the simulated measurements
like so $$\bar{o}_t = \frac{1}{B} \sum_{b=b_1}^{b_B} b \cdot f_t[b].$$

Group polarization is the difference between the post-deliberation mean observed opinion and the
pre-deliberation one, which we can write as $$g = \bar{o}_\mathrm{post} - \bar{o}_\mathrm{pre}.
  \label{eq:group-polarization}$$

@@

### Simulated spurious group polarization

Group polarization measures radicalization in experimental echo chambers in terms of the difference
between mean group opinions before and after the group discusses some survey topic. The following
model shows the problem of how the measurement scheme above can make *mere agreement* spuriously
look like polarization.

To make things more exact, we define a *null model*, $\nu$, of mere agreement defined as a 3-tuple:
a constant mean for the pre- and post-discussion latent opinion distributions, and a
pre-deliberation latent opinion variance that is greater than the post-deliberation variance to
represent finding agreement, $$
\nu = (\mu,~\sigmapre,~\sigmapost).
$$ Group polarization is *spurious* in this case if measurement clipping causes the observed
difference in pre- and post-discussion opinions to be greater than zero, i.e., $g > 0$ (Equation
[\[eq:group-polarization\]](#eq:group-polarization){reference-type="ref"
reference="eq:group-polarization"}, even if the true latent change is zero.

We will show how this arrangement generates spurious polarization in an experimental condition in
our corpus, the *Americans* experimental condition in @Moscovici1969, to which we gave the
Experiment ID `Moscovici1969_Americans`. In that condition, the authors used a seven-bin scale
($B=7$) with numerical values from -3 to +3 indicating opinions from "strongly disagree" to
"strongly agree". Pre-consensus opinions were reported to have mean -0.6 and the post-consensus mean
was -1.04 [@Moscovici1969 p. 132, Table 4].

And now we will show that a null model with
$\nu = (\mu = -1.2, \sigmapre = 4.3, and \sigmapost = 1.9)$ generates spurious polarization with
exactly the pre- and post-discussion means Moscovici and Zavalloni report
(Figure [\[fig:distros\]](#fig:distros){reference-type="ref" reference="fig:distros"}). First,
calculate the vector of probability density values for each bin, $f[b]$, for the pre- and
post-discussion latent opinion distributions using Equation [\[eq:\]](#eq:){reference-type="ref"
reference="eq:"} (calculated with
\['makeProbVec'\](https://github.com/mt-digital/gp-statmod/blob/final-draft/GroupPolarizationStatmod/model.R#L43)
from the \[project code
repository\](https://github.com/mt-digital/gp-statmod/tree/final-draft/GroupPolarizationStatmod)

$$
    f_\mathrm{pre}[b] = 
$$ $$
    f_\mathrm{post}[b] = 
$$

::: {style="text-align: center;"}
FIGURE [\[fig:distros\]](#fig:distros){reference-type="ref" reference="fig:distros"} ABOUT HERE
:::

## Root-finding Algorithm to Find Null-Polarization Models with Artifacts

>>>>

---
editor: 
  markdown: 
    wrap: 80
bibliography: this.bib
---

\newcommand{\mupre}{\mu_\mathrm{pre}}
\newcommand{\mupost}{\mu_\mathrm{post}}
\newcommand{\sigmapre}{\sigma_\mathrm{pre}}
\newcommand{\sigmapost}{\sigma_\mathrm{post}}

\newcommand{\thetafit}{\theta^\mathrm{fit}}

\newcommand{\fdr}{\mathrm{FDR}}

\newcommand{\meanobs}{\bar{o}}
\newcommand{\meanobsemp}{\bar{\mathbf{o}}}
\newcommand{\meanobst}{\meanobs_t}
\newcommand{\meanobsjt}{\meanobs_{j,t}}
\newcommand{\meanobspre}{\meanobs_\mathrm{pre}}
\newcommand{\meanobspost}{\meanobs_\mathrm{post}}
\newcommand{\tpre}{t_\mathrm{pre}}
\newcommand{\tpost}{t_\mathrm{post}}

@@

A group polarization replication is proven undecidable, and therefore not truly
a replication, if we can find a model of *mere agreement* that appears to be
polarized under that replication's experimental design. If we cannot find such a
model, we cannot determine a replication to be undecidable. Mere agreement in
this model is a dynamic where agreement increases within the group, but
extremism does not—in other words, A *null-polarization distributions* is a set
of two Gaussian distributions: one representing the pre-discussion opinions of
the group, and the other representing the post-discussion opinions. Our
*null-polarization distributions* represent the process of the group finding
*mere agreement*, which is when the average group opinion is constant over time,
but opinion variance decreases to represent agreement (or, equivalently,
*consensus* [@DeGroot1974]).

@@

## Simulation of Mere Consensus and Opinion Change Measurement

We model \emph{mere consensus} and opinion change measurement in group
polarization experiments to examine whether published replications can be
explained by this particular null polarization model. We simulate consensus and
measurement two ways for complementary purposes. First is this *expectation*
model, and the second is the *stochastic* model. Both models assume group
opinions are normally distributed, which is practical due to its simplicity.

We assume that participants' true latent opinions at some time, $t$, are random
variables drawn from a normal distribution with mean $\mu_t$ and standard
deviation $\sigma_t$. A normal distribution is scientifically justifiable, since
opinions are aggregations of many constituent electrical and chemical signals in
the brain—the distribution of aggregations like this tend to be *normal* (hence
the name). A normal distribution also makes sense because several of the
analysis procedures from the corpus use a $t$-test or other basic linear
regression-style test, which implicitly assume data are distributed normally.

The two models differ in how they simulate opinions, opinion change, and opinion
change measurement. We use the *expectation model* as a computationally
efficient way to simulate the observed mean opinion as the mathematical
*expectation value* of the distribution of measured opinions. In the second
model, the *stochastic model*, we simulate participant survey reporting as
drawing a latent opinion from the latent opinion distribution, then binning that
opinion in the associated opinion bin. We use the expectation model for its
efficiency in the first step of our analysis: to find a *null-polarization
model* that also fits the data for each experimental condition studied—if we
find one that means we must retract this replication since this proves that the
group polarization is undecidable with the data we have. Once we find any such
undecidable results, we use the null-polarization model to help us further
evaluate the experimental design itself.

In the *stochastic model*, we simulate group polarization data for the
*Evaluation* step of the analysis. We simulate experimental data, with the same
number of participants as the original experiment, using the null-polarization
model found in the *Identification* step, if one was found.

We *evaluate* the experimental design, $e$, in terms of the false positive rate,
also known as the Type I error rate or family-wise error rate, denoted
$\alpha_e$. We close our analysis by demonstrating how our method can be
inverted to *evaluate* experimental designs in the pre-registration stage to
avoid preventable replication failures like we identify here.

### Latent Opinion Model

Throughout this paper we model a participant $i$'s latent opinion at time $t$ as
a normally-distributed random variable,

$$ 
\omega_{i,t} \sim
\mathcal{N}(\mu_t, \sigma_t).  
$$
Group polarization in latent opinions is defined as any pair of pre- and
post-discussion opinion distributions where the mean increases in magnitude over
time—formally this is $| \mupre | < | \mupost |$. We need to take magnitudes since radicalization may be towards increasingly negative or positive opinion values.

For group polarization experiments, only two times are relevant:
*pre-discussion*, $t=\mathrm{pre}$, and *post-discussion*, $t=\mathrm{post}$. Since participant groups are assumed to be likeminded, we could also assume they come to agree more over time as well as become more extreme in our group polarization model, or in terms of the latent opinion variance, $\sigmapre > \sigmapost$—but only one study in our ten actually measured opinion variance to check this [@Schkade2010].

A *null-polarization hypothesis*, or *model*, is one where the latent mean stays
constant, i.e. $\mu = \mu_\mathrm{pre} = \mu_\mathrm{post}$. *Mere agreement* is
one type of null-polarization model, 
the variance decreases over time to model agreement as described above. The mere agreement model is the cornerstone of our analysis, since it is empirically plausible and generates *spurious* observations of group polarization under mere agreement models, due to measurement artifacts induced by clipping all latent opinion values outside the measurement range—we call this *spurious polarization* for short. We simulate opinion measurements in two different ways, but both use the mere agreement model of latent opinions to test whether and by how much the measurement procedure distorts null-polarization to look like group polarization.

@@

### The Expectation Model of Opinion Measurement

We use the \emph{expectation model} of opinion measurement to \emph{Identify} which, if any, mere agreement null-polarization models generate spurious group polarization in a particular experiment, $e$. The trick to this model, and the source of its name, is that we equate the theoretical \emph{expected value} of opinion measurements with the simulated \emph{observed average} of participant opinions.

We simulate the measurement of the mean observed opinion at time $t$ 
by first calculating integrating the normal probability density function over measurement bin boundaries $f(\omega)$, over each bin's corresponding range of opinion values. For example, consider an ordinal scale  that runs from -3 (strongly disagree) to +3 (strongly agree) (as
used by @Moscovici1969, for example). the frequency with which the "2" bin is
chosen is just the total area underneath the probability density curve between
1.5 and 2.5.

The probability density that a participant reports the opinion in bin index $i$ at time $t$ is
$$
f_{it}(\mu, \sigma_t) = \int_{\theta_{i - 1}}^{\theta_{i}} f(\omega; \mu, \sigma_t)~d\omega,
$$
which we calculated in R to evaluate the integral at the two threshold values $\theta_{i-1}$ and $\theta_i$. Note that the normal probability density function we used is parameterized by mean $\mu$ and standard deviation $\sigma_t$.
$$
f(\omega; \mu, \sigma_t) = 
    \frac{1}{\sigma_t \sqrt{2 \pi}} e^{-\frac{(\omega - \mu)^2}{2 \sigma_t^2}}.
$$

To simulate the measurement of the *average* observed opinion at time $t$, $\bar{o}_t$, as the expected value of the probability density function of observed ordinal opinions. In other words, we simulate $\bar{o}_t$ by calculating the weighted sum of products between the probability of observing each bin value and the bin value itself:
$$
\bar{o}_t(\nu, e) = \sum_{i=1}^B b_i f_{it}(\nu, e).
$${#eq-sim-ave-observed}

Group polarization is the difference between the post-deliberation mean observed
opinion and the pre-deliberation one. We will calculate the magnitude of spurious group polarization under a mere agreement null model represented by its collected distribution parameters, $\nu = (\mu, \sigmapre, \sigmapost)$, under experimental design, $e$: 
$$
g(\nu, e) = |\bar{o}_\mathrm{post}(\nu, e)| - |\bar{o}_\mathrm{pre}(\nu, e)|
$${#eq-spurious-gp}

In the *Inspect* stage of our analysis we will determine which group polarization replications are invalidated due to equifinality between the *mere agreement null-polarization hypothesis* on the one hand, and the *group polarization hypothesis*
on the other. This requires more than just showing there is some $\nu$ that causes spurious polarization for some experiment where $g(\nu, e) > 0$. We have to find a $\nu$ that generates the exact $g(\nu, e)$ reported in the original journal article *and* check that simulated $\bar{o}_\mathrm{pre}$ and $\bar{o}_\mathrm{pre}$ match published values.

@@

@@

#### Example: the "Americans" condition in Moscovici and Zavalloni (1969)

One of the replications we invalidate here is from Moscovici and Zavalloni (1969),
the "Americans" condition as they called it—we coded it `Moscovici1969_Americans`. 
They report that participant opinions polarized after discussing the survey topic, "American economic aid is always used for political pressure." They report survey 
responses increased in extremity from an average of -0.9 before discussion to
-1.6 after, on a seven-bin scale
from -3 indicating "strongly disagree" to +3 indicating "strongly agree".

Moscovici and Zavalloni say their results confirm the basic hypothesis, that \emph{group discussion causes polarization}. But we found a mere agreement null-polarization model that generates spurious group polarization using our model—specifically, $\nu = (\mu = -1.2, \sigma_\mathrm{pre}= 4.3, and \sigma_\mathrm{post}= 1.9)$. This means that the hypothesis is actually undecidable given the data, invalidating Moscovici and Zavalloni's replication.
When a hypothesis and its null explain data equally well, the data tell us nothing.

Now let's use the expectation model to prove that this $\nu$ generates spurious group polarization in this experiment, i.e., $g(\nu, e= \mathrm{Moscovici1969\_Americans}) > 0$. 
In the process we demonstrate how the expectation model works to simulate the measurement of spurious group polarization (@eq-spurious-gp).

First, we calculate the probability density vector using the [`makeProbVec`](https://github.com/mt-digital/gp-statmod/blob/main/GroupPolarizationStatmod/model.R#L44) function in the file [`GroupPolarizationStatmod/model.R`](https://github.com/mt-digital/gp-statmod/blob/main/GroupPolarizationStatmod/model.R) in the [project code repository on GitHub](https://github.com/mt-digital/gp-statmod) for both experiment times: pre- and post-discussion:
$$
f_\mathrm{pre} = 
\begin{pmatrix}
  0.38 & 0.09 & 0.09 & 0.09 & 0.08 & 0.07 & 0.19
\end{pmatrix}
$$ 
and
$$
f_\mathrm{post} = \
\begin{pmatrix} 
0.25 & 0.19 & 0.21 & 0.17 & 0.11 & 0.05 & 0.03 
\end{pmatrix}.
$$

When we plug in the bin values and copmuted $f_\mathrm{pre}$ values into the simulated average pre-discussion opinion definition (@eq-sim-ave-observed),
$$
\bar{o}_\mathrm{pre} = -3 \cdot 0.38~~+-2\cdot 0.09~~+\ldots+3\cdot0.19 = -0.6
$$
$$
\bar{o}_\mathrm{post} = -3 \cdot 0.25~~+-2\cdot 0.19~~+\ldots+3\cdot0.03 = -1.04
$$
So, we find that for the `Moscovici1969_Americans`, our null-polarization model appears polarized, with the exact same average pre- and post-discussion opinions reported by Moscovici and Zavalloni (1969), and the same magnitude of radicalization (CHECK THE PRESENTATION OF ALL THREE ABOVE)
$$
g(\nu, e) = \bar{o}_\mathrm{post} - \bar{o}_\mathrm{pre} = -1.04 - (-0.6) = - 0.4.
$$
::: {style="text-align: center;"}
FIGURE [\[fig:distros\]](#fig:distros){reference-type="ref"
reference="fig:distros"} ABOUT HERE 
:::

The replication of the `Moscovici1969_Americans` experimental condition is therefore invalidated since we proved equifinality between the group polarization hypothesis 
(discussion induces polarization) and its null (discussion does not induce polarization). \emph{If the null fits, we must omit} the result from the set of results supporting any particular hypothesis. Therefore, despite being cited over 1,000 times as a replication, these citations are in fact erroneous—even if powerful authors hold them up as such.

But how did we actually \emph{find} the null-polarization model that generated spurious group polarization? Below we demonstrate how we use a stochastic descent-style root finding that calculates spurious group polarization for a given $\nu$, then takes a step in a random direction, biased towards the direction of less error. It is therefore essential to use this efficient calculation simulation method, since tens, hundreds, or more evaluations are required for this root-finding algorithm. We show exactly how we put this to work in the following subsection explaining the algorithm to finish the \emph{Identification} step of our analysis.

<!-- OLD STUFF BELOW BEFORE QMD REWRITE -->
<!-- Moscovici and Zavalloni report -->
<!-- (Figure [\[fig:distros\]](#fig:distros){reference-type="ref" -->
<!-- reference="fig:distros"}). First, calculate the vector of probability density -->
<!-- values for each bin, $f[b]$, for the pre- and post-discussion latent opinion -->
<!-- distributions using Equation [\[eq:\]](#eq:){reference-type="ref" -->
<!-- reference="eq:"} (calculated with -->
<!-- \['makeProbVec'\](https://github.com/mt-digital/gp-statmod/blob/final-draft/GroupPolarizationStatmod/model.R#L43) -->
<!-- from the \[project code -->
<!-- repository\](https://github.com/mt-digital/gp-statmod/tree/final-draft/GroupPolarizationStatmod) -->
## Root-finding Algorithm to Find Null-Polarization Models with Artifacts
